---
title: "Interpolation" 
subtitle: "Theory, Mathematics, and Implementation using DemogAnr package"
author: "Edward Owusu Manu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpolation: Theory, Mathematics, and Implementation in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

Interpolation is a mathematical technique used to estimate unknown values within the range of known data points. It plays a crucial role in demographic analysis, allowing researchers to estimate missing values, smooth population trends, and project demographic indicators.

In demographic studies, data is often collected at discrete time points, leaving gaps that need to be filled for accurate analysis and decision-making. Interpolation provides a systematic way to estimate values between observed data points, ensuring a more comprehensive understanding of trends over time. This technique is particularly useful for handling population estimates, fertility rates, mortality rates, and other demographic indicators.

There are several types of interpolation methods, each with its own strengths and use cases. This vignette covers three key methods:

-   **Lagrange Interpolation**: A polynomial interpolation method that constructs a single polynomial passing through all given points. It is useful for small datasets but can be computationally intensive for larger datasets.
-   **Newton’s Interpolation**: A method that builds the interpolating polynomial incrementally using divided differences, making it more efficient than Lagrange interpolation.
-   **Spline Interpolation**: A piecewise polynomial interpolation that ensures smooth transitions between data points, widely used for demographic data due to its flexibility and stability.

This vignette explores these interpolation methods in detail, explaining their mathematical foundations and demonstrating their implementation using the `DemogAnr` package in R.

# Mathematical Formulation

## Polynomial Interpolation

A polynomial interpolation constructs a polynomial of degree $n-1$ that passes through $n$ given points:

$$
P_n(x) = a_0 + a_1x + a_2x^2 + \dots + a_{n-1}x^{n-1}
$$

where the coefficients $a_i$ are determined by solving a system of equations. Polynomial interpolation is useful when data points are smooth but can become unstable for large $n$.

## Lagrange Interpolation

Lagrange interpolation constructs an interpolating polynomial as:

$$
P_n(x) = \sum_{i=1}^{n} y_i L_i(x)
$$

where the Lagrange basis polynomials are:

$$
L_i(x) = \prod_{\substack{j=1 \\ j\neq i}}^{n} \frac{x - x_j}{x_i - x_j}
$$

This method avoids solving linear systems but can be computationally expensive for large datasets.

## Spline Interpolation

Spline interpolation fits piecewise polynomials to the data points, ensuring smooth transitions. A **cubic spline** is given by:

$$
S_i(x) = a_i + b_i(x - x_i) + c_i(x - x_i)^2 + d_i(x - x_i)^3
$$

with continuity constraints on the first and second derivatives:

$$
S_i'(x_i) = S_{i+1}'(x_i), \quad S_i''(x_i) = S_{i+1}''(x_i)
$$

Spline interpolation is widely used in demographic analysis due to its flexibility and smooth results.

# Implementation in R

## Load the DemogAnr package

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
library(DemogAnr)
```

## Example 1: Interpolating between points (vector input)

Before running the interpolation, we define a simple dataset:

```{r}
x <- c(1990, 2000, 2010, 2020)
y <- c(24.5, 26.3, 28.1, 29.8)  # Example: Population in millions
x_new <- c(1995, 2005, 2015)

# Perform interpolation
interpolation(x, y, x_new, method = "linear")
```

## Example 2: Interpolating for a dataframe

Let's assume we have a dataset of demographic indicators:

```{r}
data <- data.frame(
  year = c(1990, 2000, 2010, 2020),
"x1990" = c(24.5, 26.3, 28.1, 29.8),
"x2000" = c(30.2, 31.1, 29.3, 33.1),
"x2010" = c(28.2, 32.3, 31.1, 34.2),
"x2020" = c(30.3, 34.1, 33.8, 35.2)
)
x_new <- c(1995, 2005, 2015)

# Perform spline interpolation
interpolation(x = data$year, data = data, y_cols = c("x1990","x2000","x2010","x2020"), x_new = x_new, method = "spline")
```

# Conclusion

Interpolation is a powerful tool for estimating missing demographic values and smoothing population trends. In this vignette, we introduced key interpolation methods, explained their mathematical foundations, and demonstrated their application using the `DemogAnr` package. Future improvements could include adaptive interpolation techniques for handling irregularly spaced demographic data.
